A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-06 latex/main.tex

\section*{DISCUSSION}
There are two major contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025on,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements which fully integrate
recombination into pathogen genomics.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.

The second major contribution is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
By capturing recombination in real time at a global scale, 
our approach extends the established utility of comprehensive 
genealogies~\cite{Hinrichs2024,McBroome2024}, 
providing clear insight and enabling timely investigation of 
recombination-aware pathogen dynamics.
The novelty of the inferred ARG motivates our detailed 
analyses of the methodology behind sc2ts, which we use to validate the ARG itself. 
This ensures that the ARG is a trustworthy basis for future research, 
deepens our understanding of the inference method, and 
suggests avenues for improvement (see Limitations of the Study).

Study of the sc2ts ARG gives insight into the evolution of SARS-CoV-2, 
which provides several further contributions to our understanding of the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Second, we use the ARG to study breakpoint intervals 
for recombination events, add a novel temporal dimension
and show a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we  provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
Together, these  underscore the utility of the ARG itself, and suggest
several important directions for future research.

% Old discussions(s)

% % P1. The ARG. Unique resource. Emphasise utility of UShER and how 
% % this has more.
% % SHZ: This makes it sound like we are just making analyses
% % (which people can do already using existing methods and tools)
% % easier to do
% % rather than making a leap forward from anything else out there.
% We present the first pandemic-scale ancestral recombination graph (ARG) for
% SARS-CoV-2, reconstructed from 2.48 million genomes using \texttt{sc2ts}. This
% ARG represents a step change in how the evolutionary history of this virus can
% be represented and analysed. Like the UShER phylogeny, it provides a
% comprehensive and continuously updatable record of global genomic diversity,
% but it extends this paradigm by embedding recombination directly into the
% genealogy. The ARG unifies point mutations, deletions, and recombination events
% within a single, compact data structure, enabling analyses that
% were previously fragmented across multiple methods and data models. By encoding
% reticulate ancestry explicitly rather than assuming a single tree, the ARG
% provides a complete and internally consistent representation of SARS-CoV-2
% evolution. In this way, the
% ARG offers the same accessibility and scalability that made the UShER tree
% indispensable during the pandemic, while expanding its scope to encompass the
% full spectrum of evolutionary processes shaping SARS-CoV-2.

% % P2. This is a conservative lower-bound on recombination. 
% % Discuss other estimates.
% % 
% % SHZ: Why should we care about getting this number/frequency right?
% % Sure, popgen folks love estimating this, but to other people,
% % existing estimates already tell us that
% % detectable recombination in SARS-CoV-2 is (qualitatively) rare,
% % even without the advances of sc2ts.
% % Also, we are talking about the frequency of detectable recombination
% % in the last sentence, right?
% % The true frequencey of recombination is unknowable from this type of data.
% % So, the best we are ever going to get is a better lower bound.
% The recombination frequency estimated from the \texttt{sc2ts} ARG is
% substantially lower than that implied by the combination of UShER+RIPPLES,
% largely reflecting differences in model formulation and parameter choice.
% RIPPLES, applied post-hoc to UShER trees of 1.6 million genomes up to May 2021,
% reported 589 unique recombination events with 43,104 descendant
% samples---corresponding to roughly 2.7\% of all genomes being of recombinant
% origin.  Using the same time interval, our inference at \(k=4\) identifies only
% 14 high-confidence recombination events with 172 descendants among 501,235
% genomes (0.034\%), a 40- to 80-fold reduction.  This difference is largely
% explained by the stricter parsimony penalty adopted here: increasing the
% penalty from \(p=3\) (as used in the RIPPLES analysis) 
% to \(k=4\) greatly reduces false
% positives while retaining nearly all verified recombinant lineages, 
% including those missed by RIPPLES at equivalent thresholds.  
% We argue that the post-hoc nature of UShER+RIPPLES can erase evidence 
% of recombination as
% tree rearrangements proceed, thereby requiring the more lenient 
% threshold in order to find known recombinats.
% These comparisons suggest that our estimate represents a conservative 
% lower bound. Further work will be required to determine
% the true frequency of recombination in SARS-CoV-2 evolution.

% % P3. Clarifies relationships between pango X lineages, simplifying
% % the Omicron sublineage.
% % 
% % SHZ: I don't think knowing the relationships between Pango Xs is what's really important.
% % Sure, phylogeneticists would get a kick out of it,
% % but I don't think the majority of Cell readership would.
% % I think what is important is knowing
% % how many emerging recombinants carrying new combinations of mutations of concern
% % we are actually dealing with.
% % Clarifying the relationships between Pango Xs helps with this task.

% The Pango X lineage designations represent the community’s best attempt to
% catalogue recombinant SARS-CoV-2 lineages, but they arise from a largely
% manual, volunteer-driven process that synthesizes multiple lines of evidence
% and expert judgment rather than a single automated criterion.  As a result, the
% system provides the closest available approximation to “truth data,” but
% inevitably contains inconsistencies and over-splitting.  Within the
% \texttt{sc2ts} ARG, 44 Pango X lineages are represented, of which 16 correspond
% directly to recombination nodes (Type I events).  Examination of their
% genealogical relationships shows that many of these 44 lineages share common
% ancestral recombination events rather than arising independently.  In
% particular, five lineages—XAA, XAG, XAM, XR, and XU—descend from the same
% recombinant ancestor associated with XQ, while another five—XZ, XAC, XAD, XAE,
% and XAP—share a second common ancestor that we refer to as “Xx.”  These
% clusters account for roughly half of all X designations in the dataset,
% indicating that the community process has at times produced multiple labels for
% the descendants of a single recombinant lineage.  Conversely, several [TODO 
% be more concrete]
% designated X lineages show little or no evidence of recombination in the ARG,
% suggesting that some may reflect convergent mutation or limited sampling rather
% than hybrid origin.  Together, these results provide a data-driven synthesis in
% which the 44 currently recognized recombinant lineages collapse into a smaller
% number of well-supported ancestral events, simplifying the complex picture of
% Omicron-era evolution and bringing the community nomenclature into closer
% alignment with the underlying genealogy.

% % P4. Results on global case numbers and frequency of recomb.
% % SHZ: Should we mention how the reliability of global case number data
% % and variation in sampling effort over time and across countries
% % might affect our interpretation of these findings?
% [TODO: This is direct output from ChatGPT.  Review!]
% The dated ARG allows us to place recombination within its epidemiological
% context, revealing clear temporal structure in when recombinant lineages arise.
% The frequency of inferred recombination events follows global case numbers
% through the pandemic, with peaks during the Delta and Omicron waves when both
% infection prevalence and genetic diversity were high.  Quantitatively, the best
% single predictor of weekly recombinant counts is the product of global cases
% and lineage heterozygosity (\(nH\)), which explains nearly a quarter of the
% variation in event frequency (\(R^{2}_{\mathrm{adj}} = 0.24\)).  This
% relationship is consistent with the mechanistic expectation that recombination
% requires both frequent co-infections (driven by high case numbers) and
% genetically distinct co-infecting viruses (driven by lineage diversity).  In
% other words, recombination becomes most detectable when epidemiological and
% evolutionary opportunity coincide.  These results imply that the emergence of
% novel recombinants is not uniform through time but concentrated into
% predictable ``hot periods'' during large, antigenically diverse outbreaks, when
% the ecological conditions for viral recombination are most favourable.

% % P5. Confirm breakpoints around Spike. 
% % SHZ: Maybe we can say it's a known (instead of robust)
% % pattern of detectable recombination in CoVs in general?
% % There are some citations for template switching hotspots in SC2.
% % Also, it's not only about selection favoring new combinations of mutations in Spike.
% % It's about combinations of mutations in, say, ORF1ab, getting selected against
% % (or just low mutation rates).
% % Maybe mention epistasis?
% [TODO: This is directish output from ChatGPT. Review!]
% The genomic distribution of breakpoints in the ARG is highly uneven, with a
% pronounced enrichment near the $5'$ boundary of the Spike gene and across the
% S1–S2 junction.  This pattern closely mirrors that observed in the
% UShER+RIPPLES dataset \citep{Turakhia2022} and in earlier 
% analyses \citep{Jackson2021} [more cites], but here it is recovered
% independently over a much longer sampling period and under a more conservative
% inference regime.  Under our more conservative estimate of the overall
% prevalence of recombination, we find that this enrichment persists despite a substantial
% reduction in the total number of inferred recombination events, indicating that
% it reflects a genuine and stable feature of SARS-CoV-2 evolution rather than a
% by-product of detection sensitivity.  Across all high-confidence events, the
% median breakpoint interval length is $\sim 2.2$kb and declines with parental
% divergence, consistent with expectations from sequence-based localization
% accuracy.  These results show that even with a conservative lower bound on
% recombination frequency, the concentration of breakpoints around Spike remains
% a robust signature of coronavirus genome biology—likely reflecting a
% combination of mechanistic biases in template switching and selective
% constraints on Spike function.

% % P6. Sc2ts method is a major advance. We validate with real data
% % rather than simulations. Forward looking statement about simulations.
% % 
% % SHZ: Could at least some of this go into the Introduction?
% In the broader context of pathogen genomics, \texttt{sc2ts} demonstrates that
% it is now possible to reconstruct recombination-aware genealogies at the scale
% of entire pandemics.  Although recombination is recognised as a major
% evolutionary force in many pathogens—including RNA viruses such as influenza
% and HIV, and numerous bacterial species—genealogical analyses have typically
% relied on tree-based approximations that ignore or collapse recombination.
% This simplification has limited the accuracy of downstream evolutionary and
% epidemiological inference, even in well-studied systems.  \texttt{Sc2ts}
% overcomes this barrier by directly inferring an ancestral recombination graph
% from millions of viral genomes, producing the first coherent reticulate
% genealogy for any pathogen dataset of comparable size.  The ability to
% integrate recombination into genome-wide genealogies at this scale represents a
% major conceptual step forward: it transforms recombination from a confounding
% complication into an analyzable component of pathogen evolution.

% % P7. The software ecosystem enables diverse besplike analysis.

% [TODO: This is close, needs revision]
% The practical value of the ARG is amplified by its integration into the broader
% software ecosystems built around \texttt{tskit} and VCF~Zarr.  These mature,
% open-source libraries provide highly efficient storage and analysis of tree
% sequences and variant data, allowing millions of genomes to be explored
% interactively within familiar environments such as Python and Jupyter
% notebooks.  Analyses that previously required bespoke phylogenetic
% pipelines—such as extracting local genealogies, summarising ancestral
% relationships, mapping mutations to branches, or tracking recombinant ancestry
% through time—can now be performed with concise, reproducible code on standard
% hardware.  Because \texttt{tskit} and VCF~Zarr share a common data model, the
% ARG integrates seamlessly with widely used tools in the PyData ecosystem,
% enabling users to combine genealogical information with epidemiological,
% geographic, or phenotypic metadata in the same workflow.  This interoperability
% transforms what can be done with large viral genome collections, lowering the
% barrier to sophisticated evolutionary analysis and making recombination-aware
% genealogy inference accessible to a wide community of researchers.

% % P8. Outlook

% [TODO a bit muddled, refine]
% Together, these results demonstrate that recombination in SARS-CoV-2 can be
% systematically incorporated into genealogical inference at pandemic scale. The
% \texttt{sc2ts} ARG captures the complexity of viral evolution in a form that is
% both biologically realistic and computationally tractable, showing that
% recombination can be modelled rather than ignored. Embedded within an
% interoperable software ecosystem, it provides a comprehensive record of
% SARS-CoV-2 ancestry and a practical framework for real-time,
% recombination-aware genomic surveillance—turning large sequence archives into
% dynamic evolutionary models that can be queried with concise, reproducible code. 
% More broadly, this work marks a transition from
% tree-based pandemic phylogenetics to recombination-aware genealogy inference.
% While our focus is SARS-CoV-2, the approach offers a blueprint and a starting
% point for other recombining pathogens: the core ideas are portable,
% even though effective deployment will require pathogen-specific adjustments. 
% We anticipate that this resource will clarify the recent history of
% SARS-CoV-2 and catalyse methodological refinement and targeted biological
% studies of recombination across systems where such extensions prove feasible.

% [NOTE: This is Shing's WIP alternative]

% %%% SHZ: WIP alternative.
% We present the first pandemic-scale
% ancestral recombination graph (ARG) for SARS-CoV-2,
% reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
% We demonstrate how this ARG, for the first time,
% enables systematic identification and characterization of
% recombination events in fine-grained
% detail--in terms of their parents, breakpoints, and timing of
% emergence--providing richer and more nuanced
% genealogical, genomic, and epidemiological context
% than existing phylogenetic approaches.
% Together, sc2ts and the ARG represent a new paradigm in
% how we study the evolution of pathogens in the face of ongoing sampling.

% Accounting for recombination is a key challenge in pathogen genomics, and
% recombination has been largely ignored in analysis of viral genomes.
% Existing phylogenetic approaches for pathogens detect recombination
% \textit{post hoc} (e.g., UShER+RIPPLES),
% treat it as a nuance (e.g., XX), or
% discard its signals from downstream analyses (e.g., XX).
% Sc2ts addresses this challenge
% by using the ARG as the underlying data model,
% explicitly representing recombination events as individual nodes and
% relating them to specific parents in the ARG.
% The ARG provides a coherent framework to determine and represent
% the relationships between recombinants and their specific parents
% (which are not possible using a tree-based data model, e.g., the MAT format of UShER).
% The ARG reveals and clarifies the relationships between
% individual samples of the same recombinant origin,
% avoiding issues such as over-naming recombinants
% (e.g., the samples assigned to
% five differently labeled recombinant Pango lineages
% descending from the “Xx” recombination event in the ARG).
% Our results show how an ARG can considerably simplify
% efforts to identify and monitor emerging recombinants,
% facilitating communication of genomic surveillance results.

% % SHZ: Very WIP
% % Discuss the new preprint from Turakhia.
% The timing of recombination events is of key interest in
% pathogen genomic epidemiology.
% because this information may yield insights into
% how the emergence of a new recombinant might be
% linked with changes in public health policies.

% The overall frequency of detectable recombination events
% estimated from the sc2ts ARG is substantially lower than
% that from UShER+RIPPLES,
% largely reflecting differences in model formulation and parameter choice.
% When applied \textit{post hoc} to an UShER phylogeny of
% $\sim$1.6 million genomes sampled up to May, 2021,
% RIPPLES reported 589 unique recombination events with 43,104 descendant
% samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
% Using the same time interval, sc2ts \(k=4\) identifies
% 14 high-confidence recombination events with
% 172 descendant samples among 501,235 samples \(0.034\%\),
% a 40- to 80-fold reduction.
% Much of this difference is explained by
% the stricter parsimony penalty used here:
% increasing the penalty from p=3 (as used in the RIPPLES analysis) to
% k=4 greatly reduces false positives
% while retaining nearly all verified recombinant lineages,
% including those missed by RIPPLES at equivalent thresholds.
% These indirect comparisons suggest that
% our estimate represents a conservative lower bound.
% In future studies, dissecting the methodological differences
% between sc2ts and UShER+RIPPLES may
% inspire new ideas to develop or improve ARG building methods
% for SARS-CoV-2 (and potentially other pathogens).

% The ARG itself is a powerful resource
% to support future studies of SARS-CoV-2.
% Our results show that the ARG captures and unifies
% important features of SARS-CoV-2 evolution,
% which are shown in different studies
% using various methods, tools, and approaches.
% The ARG also compactly encodes genetic variation in
% millions of SARS-CoV-2 genomes,
% allowing for convenient and efficient reuse of
% this valuable dataset for a range of
% genetic and evolutionary analyses beyond those presented here,
% for instance, to understand
% how the variability in sequencing efforts
% across time and location affects our ability
% to reconstruct the origin of important recombinant lineages and
% to estimate when they emerge.
% Because building the ARG is computationally expensive, and
% checking and improving various aspects of its quality is laborious,
% we hope that making this resource openly available will
% save significant time and effort for other researchers,
% enabling immediate focus on analyzing the information
% contained within the COVID-19 dataset.

% % TODO: Briefly discussion software tools here or in a subsequent paragraph.

% Recent developments in ARG methods have focused almost exclusively on
% applications to human sequencing data.
% The availability of human ARGs reconstructed from large datasets
% has already had a tremendous impact on improving
% our understanding of human genetic ancestry and past demography.
% We anticipate similar advances to be made
% by adapting existing inference methods for human ARGs
% to use the sc2ts ARG for the genealogy-based study of
% natural selection or the geographic spread of SARS-CoV-2 through time.

% The unprecedented throughput and pace of genome sequencing are
% creating ample opportunity to understand the impacts of recombination
% on pathogen evolution and epidemiology.
% We hope that this study will spur
% development of ARG methods for pathogens,
% ultimately making recombination-aware genome analysis
% routine for all sorts of pathogens beyond SARS-CoV-2.

@@

We have focused on applying sc2ts to the Viridian dataset rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to deal with the various data quality issues.
Secondly, the Viridian data is publicly available and redistributable,
enabling us to share the resulting ARG without restrictions.

% Need to be careful here to list stuff that sounds interesting and 
% hard enough to not get asked to do it.
The sc2ts method may be extended and improved in several directions.
Firstly, the simplistic parameterisation of the LS HMM used 
here could be extended to include more realistic
models with mutation rate variation along the genome~\cite{Haddox2025}
and transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Secondly, the current handling of deletions by post-hoc parsimony
could be improved either by a more nuanced data-masking strategy
(in which, e.g., all but the right-most base of known deletions is 
masked as missing data during HMM matching), or by the more 
challenging approach of incorporating multi-base events into the HMM directly.
Sc2ts is currently specialised for SARS-CoV-2 data, but the methods
may be readily adapted to other pathogens.

We have used the tsdate method~\cite{Wohns2022,Pope2025} to date the 
sc2ts ARG, and initial results are promising. However, dating 
of phylogenies in itself is a complex task~\cite{Kumar2016advances}
and methods for incorporating information from metadata of uncertain 
quality into large trees are in their 
early stages~\cite{sanderson2021chronumental} and there is certainly
substantial room for improvement.
Much of the complexity of sc2ts lies in 
dealing with ``time traveller'' samples, where sampling dates 
provided in metadata are substantially incorrect. 
The present ARG may provide 
a way of improving this, by iteratively matching samples against
it to estimate dates for all sequences, and adjusting the dates
for time-travellers detected in this way. Sc2ts could then
be run without the complex filters required to remove 
time-travellers and unconditionally include all but the most divergent
sequences. Other studies may also benefit from the 
improved sample dates.

[Final para is prelim. Acknowledge limitations, forward looking 
for how we might deal with them]
Finally, it is important to acknowledge that sc2ts performs best
with dense sampling, and the dataset is heavily geographically biased.
[cite ~\cite{Lancet2021,Kalia2021,Chen2022}?]
Long branches, cause particular problems. Future improvements would 
ideally detect when data is sparse and switch to more sophisticated
models, as discussed in~\cite{Bisschop2025likelihoods}.

% It is challenging to accurately reconstruct global events
% when there are regions with poor sampling and
% jurisdictions with long delays
% before sequences become publicly
% available~\cite{Lancet2021,Kalia2021,Chen2022}.
% For example, revisiting the ARG topology periodically
% (e.g., rewinding the ARG a month and repeating
% with samples that became available later)
% might improve the accuracy of the genealogy.

% We have developed and used several methods
% to exclude spurious recombination events \textit{post hoc},
% including filtering by supporting loci per parent and
% splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
% These methods could instead be run during ARG inference,
% immediately after sample matching,
% reducing the number of ARG post-processing steps.

% Finally, an inherent limitation of detecting recombination from sequence data,
% which applies to all methods,
% is that it is impossible to detect events with few (or no) supporting bases,
% as these cannot be distinguished from mutation events.
% This limitation makes it particularly challenging
% to detect events near the termini of linear viral genomes.

@@

% Shifting breakpoints is a minor point, not worth mentioning?
We computed Pango assignments for all 2,747,985 nodes in the ARG by first 
exporting their aligned sequences to a 77GiB FASTA file and 
then running pangolin (version 4.3.1; pangolin-data version 1.29).
The resulting Pango and Scorpio designations were then stored as 
metadata associated with each node in the ARG.

>>>>

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware 
methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025on,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.
% NOTE: still missing the "What" bit suggested by Jotun, where we 
% list out some of the stuff that could be done. I'm not convinced
% this works, though.

The other primary contribution of this work is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
To ensure these innovations are well-founded, we have paid significant 
attention to methodological detail, carefully validating the ARG
(and consequently the method) against known results. 
Sc2ts is currently specialised to take advantage of unique features 
of the SARS-CoV-2 dataset, but the methods may be readily adapted
to other pathogens. The LS HMM is the most important element of 
sc2ts, efficiently finding parsimonious placements for samples 
among all possible recombinant solutions. The current 
parameterisation---emphasising interpretability---is simplistic,
and could readily be extended to account
for mutation rate variation along the genome~\cite{Haddox2025}
or transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Our analysis of deletions shows that valuable phylogenetic signal
is lost when they are ignored, and an important avenue for 
future work is to explore ways of incorporating them into the HMM.
Another methodological innovation is the use of 
the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
in the ARG. While results are encouraging, 
dating of phylogenies is a complex task~\cite{Kumar2016advances}
and methods for incorporating information from metadata of uncertain 
quality into large genealogies are in their 
early stages~\cite{sanderson2021chronumental}, and there is certainly
substantial room for improvement.
Sc2ts itself may benefit from these node dates, as 
much of the method's complexity lies in dealing with 
sample with incorrect sampling dates, and the ARG may provide 
a means of pre-identifying such ``time travellers''.

The secondary contributions of this work arise from 
the detailed study of the sc2ts ARG, providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X-lineages, suggesting significant 
clustering of named lineages under shared recombination events.
Second, we used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we  provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
Together these results underscore the utility of the ARG itself 
and suggest important directions for future research.

@@

We have focused on applying sc2ts to the Viridian dataset~\cite{Hunt2024} rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to address data quality issues.
Secondly, the Viridian data is publicly available and redistributable,
enabling us to share the resulting ARG without restrictions.

It is important to acknowledge that sampling is nonuniform both
geographically and over time~\cite{Lancet2021,Chen2022}. 
Sc2ts performs best in the data-dense regime, when the underlying
assumption that genetic diversity is gradually accumulating within
the structure holds. Large changes (i.e, saltation events) present
significant challenges and the results of the algorithm should 
be regarded more critically in these situations.
Future methodological improvements may combine the strengths 
of sophisticated model-based inferences with the power 
of non-parametric inference~\cite{Bisschop2025likelihoods}
when such data sparsity is detected.

@@

% Shifting breakpoints is a minor point, not worth mentioning?
We computed Pango assignments for all 2,747,985 nodes in the ARG by first 
exporting their aligned sequences to a 77GiB FASTA file and 
then running pangolin (version 4.3.1; pangolin-data version 1.29).
Note that this version of pangolin uses UShER placement
for classification~\cite{McBroome2024}.
The resulting Pango and Scorpio designations were then stored as 
metadata associated with each node in the ARG.
